---
title: "Part 2(c)"
author: "Chin Joey"
date: "2025-04-03"
output: html_document
---

Part 2(c)
1.Exclude cancelled flights
2.Select key features 
3.Convert variables to factors for modeling
4.Split data into training (40%) and testing (60%) sets
5.Create a logistic regression task for flight diversion prediction
6.Train the model and evaluate its performance
7.Analyze coefficients yearly to see feature impacts over time
8.Visualize results with a trend plot of coefficients

Load libraries
```{r}
library(data.table)
library(mlr3)
library(mlr3learners)
library(mlr3pipelines)
library(ggplot2)
library(scales)

```

Load data and import it into data frame
```{r}
airports_df <- read.csv("airports.csv", header = TRUE)
carriers_df <- read.csv("carriers.csv", header = TRUE)
planes_df <- read.csv ("plane-data.csv", header = TRUE)
ontime_df <- data.frame()
for (year in 2004:2008) {
  file_path <- paste0(year, ".csv/", year, ".csv")
  yearly_data <- read.csv(file_path, header = TRUE)
  ontime_df <- rbind(ontime_df, yearly_data)
}
```

Convert data to a data.table and exclude cancelled flights
```{r}
# Convert ontime_df to a data.table
setDT(ontime_df) 

# Replace cancelled flights with NA and remove them
ontime_df[, Cancelled := fifelse(Cancelled == 1, NA, Cancelled)]
ontime_df <- na.omit(ontime_df, cols = "Cancelled")

```

Feature Selection
1.Keep needed features for the logistic regression model
2.'Year' for visualization
```{r}
part_c_df <- ontime_df[, .(Year, DayOfWeek, CRSDepTime, CRSArrTime, UniqueCarrier, Distance, Diverted)]


```

Converted categorical variables to factors for modeling
```{r}
part_c_df[, Year := as.factor(Year)]  # Keep Year as factor for plotting
part_c_df[, DayOfWeek := as.factor(DayOfWeek)]
part_c_df[, UniqueCarrier := as.factor(UniqueCarrier)]
part_c_df[, Diverted := as.factor(Diverted)]  # Target variable

```

Split data into training (40%) and testing (60%) sets
```{r}
set.seed(123)
n <- nrow(part_c_df) # Get total number of rows
train_set <- sample(n, round(0.4 * n))
test_set <- setdiff(1:n, train_set)

```

Set up a classification task with selected features
1.Define classification task
2.Select relevant features for the model (excluding the target variable)
3.Define classification error as the performance measure
4.Print task summary
```{r}
task <- TaskClassif$new("diversion", backend = part_c_df, target = "Diverted")
task$select(c("DayOfWeek", "CRSDepTime", "CRSArrTime", "UniqueCarrier", "Distance"))
measure <- msr("classif.ce")
task

```
Logistic regression model with preprocessing
1.Define logistic regression learner
2.Create a preprocessing pipeline
3.Combines preprocessing and model into a pipeline
```{r}
learner_lr <- lrn("classif.log_reg")
gc_lr <- po("imputemean") %>>% learner_lr
glrn_lr <- GraphLearner$new(gc_lr)

```

Trains the model on the training set
```{r}
# Train the model on the training set
glrn_lr$train(task, row_ids = train_set)

```

Evaluate the model’s performance
1.Make predictions on the test set
2.Compute classification error
```{r}
predictions <- glrn_lr$predict(task, row_ids = test_set)
predictions$score(measure)

```
Yearly Coefficient Analysis
1.Create empty list to store coefficients
2.Loop through each year 
3.Fit logistic regression for that year's data
4.Extract coefficients for key flight features
5.Filter out individual categories, keeping only main effects
6.Combine all years' results into one data frame
```{r}
coef_list <- list()

for (yr in unique(part_c_df$Year)) {
  
  year_data <- part_c_df[Year == yr] # Subset data for the specific year
  
  model <- glm(Diverted ~ CRSDepTime + CRSArrTime + Distance + UniqueCarrier + DayOfWeek,
               data = year_data, family = binomial)
  
  coef_dt <- data.table(Year = yr, Term = names(coef(model)), Coefficient = coef(model))
  
  coef_dt <- coef_dt[Term %in% c("CRSDepTime", "CRSArrTime", "Distance", "UniqueCarrier", "DayOfWeek")]

  coef_list[[as.character(yr)]] <- coef_dt # Store results
}

coef_data <- rbindlist(coef_list)

```

Plot how feature importance (coefficients) changes across years 
```{r}
ggplot(coef_data, aes(x = Year, y = Coefficient, color = Term, group = Term)) +
  geom_line() + 
  geom_point() +
  labs(title = "Logistic Regression Coefficients Over Time",
       x = "Year",
       y = "Coefficient",
       color = "Feature") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = label_number(accuracy = 0.001))  # Format y-axis with decimals
```
